---
title: "R Notebook"
output: html_notebook
---

# Wrangling flow data

While it is convenient to use a package such as `cytofkit`, sometimes you need to further manipulate your data or experiement with a new clustering or visualization algorithm. In order to do so, it is necessary to know how to manuplate data "from first principles" using modern R syntax.

#### Load FCS file into FlowFrame

While you can load FCS files into a FlowFrame using `flowCore`, it is quite strcit in data validatioan and might not work with processed FCS files. I find it easier to just use a regular data frame and manipullate it using `dplyr` and friends.

The [package description](https://www.bioconductor.org/packages/devel/bioc/vignettes/flowCore/inst/doc/HowTo-flowCore.pdf) for `flowCore` is quite old, and installation can be tricky, so we will just show a small example for completeness.

```{r}
library(flowCore)
library(flowViz)
```

```{r}
path <- "data/AMJ_Costim_1.fcs"
x <- read.FCS(path, transformation=TRUE)
summary(x)
```

```{r}
options(repr.plot.width = 4, repr.plot.height = 4)
```

---
scrolled: false


```{r}
plot(x[1:1000, c(1,4)])
```

### Load data set into R DataFrame

[Feahtr](http://blog.rstudio.com/2016/03/29/feather/) is a fast storage format for data frames that allows for convenient data exchange between R and Python `pandas`. Given the size of flow and scRNA-seq data files, it is faster than other portable formats.

```{r}
library(feather)
library(tidyverse)
```

```{r}
library(viridis)
```

```{r}
path <- 'data/flow27parameter_indexed.feather'
df <- read_feather(path)
```

#### Instpect data set

```{r}
dim(df)
```

```{r}
colnames(df)
```

#### Convert column names to be R friendly

```{r}
colnames(df) <- make.names(colnames(df))
```

```{r}
colnames(df)
```

```{r}
head(df, n=3)
```

```{r}
tail(df, n=3)
```

```{r}
sample_n(df, 3)
```

#### Piping

```{r}
df %>% head(10) %>% tail(3)
```

#### Can split over several lines for readbility

```{r}
df %>%
head(10) %>%
tail(3)
```

#### To save piped output, re-assign to new variable

```{r}
x <- df %>%
head(10) %>%
tail(3)
```

```{r}
x
```

#### Subsetting data

```{r}
df[1:3, c(3,4,6)]
```

#### Selecting columns

```{r}
df %>%
select(c('CD4_BUV661', 'CD8_PerCP.Cy55')) %>%
head(3)
```

```{r}
df %>%
select(starts_with('CD4')) %>%
head(3)
```

```{r}
df %>%
select(contains('SC')) %>%
head(3)
```

#### Filtering rows

```{r}
dim(df)
```

```{r}
colnames(df)
```

```{r}
dim(df)
```

```{r}
df %>%
filter(FSC.A > 50000) %>%
dim
```

#### DIY rectangular gating

```{r}
df %>%
filter(FSC.A > 50000,
       FSC.A < 250000,
       CD3_BV480 > quantile(CD3_BV480, 0.5)) %>%
dim
```

#### Mutating data

`mutate` returns original and transformed columns

```{r}
df %>%
select(contains('SC')) %>%
mutate(FSC.SC = scale(FSC.A),
          SSC.SC = scale(SSC.A)) %>%
head(3)
```

`transmute` only returns transformed columns

```{r}
df %>%
transmute(FSC.SC = scale(FSC.A),
          SSC.SC = scale(SSC.A)) %>%
head(3)
```

#### Summarizing data

```{r}
df %>%
summarize_if(is.numeric, median)
```

```{r}
df %>%
select(starts_with('CD4')) %>%
summarize_all(funs(min, max))
```

### Simple visualizationm

```{r}
n <- 10000
sc <- df %>%
sample_n(n, replace = FALSE) %>%
select(contains('SC'))
```

```{r}
sc %>% head(n=3)
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 4)
```

```{r}
ggplot(df) +
geom_histogram(bins = 25, aes(x=CD45RA_BB515, alpha=0.2, fill="red")) +
geom_histogram(bins = 25, aes(x=CD4_BUV661, alpha=0.2, fill="green")) +
guides(alpha=FALSE, fill=FALSE)
```

```{r}
df %>%
select(starts_with('CD1')) %>%
gather  %>%
ggplot(aes(x=value, fill=key)) +
geom_histogram(binwidth=0.5, alpha=0.5) +
guides(alpha=FALSE)
```

```{r}
options(repr.plot.width = 4, repr.plot.height = 4)
```

```{r}
ggplot(sc, aes(x=FSC.A, y=SSC.A)) +
geom_point(shape='.')
```

[Original source for get_density function](http://slowkow.com/notes/ggplot2-color-by-density/)

```{r}
get_density <- function(x, y, n = 100) {
  dens <- MASS::kde2d(x = x, y = y, n = n)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}
```

```{r}
d <- get_density(sc$FSC.A, sc$SSC.A)
```

---
scrolled: false


```{r}
ggplot(sc, aes(x=FSC.A, y=SSC.A, color=d)) +
geom_point(shape='.') +
scale_color_viridis() +
guides(color=FALSE)
```

```{r}
colnames(df)
```

```{r}
ggplot(df,
       aes(x=CD4_BUV661, y=CD8_PerCP.Cy55,
           color=df$CD3_BV480)) +
geom_point(size=0.1) +
scale_color_viridis() +
guides(color=FALSE)
```

#### Make some subsampled data and save to file

```{r}
for (i in 1:3) {
    path <- paste('data',
                      paste('xs', i, '.feather', sep=''),
                      sep='/')
    data <- sample_n(df, 10000, replace=FALSE)
    write_feather(data, path)
}
```

```{r}
list.files('data', 'xs.*feather')
```
